$R^*$: A robust MCMC convergence diagnostic with uncertainty using gradient-boosted machines

### 2020-10-09 src/s_ml_comparison.Rmd
set up infrastructure for hyperparameter runs and ml comparison

  if(!is.null(caretGrid))
    gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_,
                   tuneGrid = caretGrid)
  else
    gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_)
  plda <- predict(object=gbmFit, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$chain) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy * n_distinct(training_data$chain))
}

@@

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet")
m_res1 <- m_res %>%
  as.data.frame() %>% 
  mutate(iter=seq_along(gbm))
saveRDS(m_res1, "../data/ml_comparison.rds")
m_res1 <- readRDS("../data/ml_comparison.rds")
g <- m_res1 %>% 
  dplyr::select(-one_of("rhat")) %>% 
  melt(id.vars="iter") %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value)) +
  geom_boxplot() +
  xlab("ML model") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14))
ggsave("../output/ml_comparison_ar1.pdf", g, width = 12, height = 8)
```

@@

tibble(var1=testing_data$V6, var2=testing_data$V10, chain=plda) %>% 
  ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
  geom_jitter()
```

Cauchy model
```{r}
fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
                refresh = 0)
x <- rstan::extract(fit_nom, permuted=F)
temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
temp$a_accuracy

fit_alt1 <- stan(file = 'cauchy_alt_1.stan',
                 seed = 7878, refresh = 0)
x <- rstan::extract(fit_alt1, permuted=F)
temp1 <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
temp <- f_replicate_ml_simple_gridded(x, "rf", tibble(mtry=1))
temp$a_accuracy
```

>>>>

  if(!is.null(caretGrid))
    if(method_!="multinom")
      gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_,
                   tuneGrid = caretGrid)
    else
      gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_,
                   tuneGrid = caretGrid,
                   MaxNWts=10000
                   )

  else
    gbmFit <- train(chain ~ ., data = training_data, 
                   trControl = trainControl(method = 'none'),
                   method=method_)
  plda <- predict(object=gbmFit, newdata=testing_data)
  a_accuracy <- 
    tibble(predicted=plda, actual=testing_data$chain) %>%
    mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
    summarise(mean(correct)) %>% 
    pull()
  return(a_accuracy * n_distinct(training_data$chain))
}

@@

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "nnet", "multinom")
m_res1 <- m_res %>%
  as.data.frame() %>% 
  mutate(iter=seq_along(gbm))
saveRDS(m_res1, "../data/ml_comparison.rds")
m_res1 <- readRDS("../data/ml_comparison.rds")
g <- m_res1 %>% 
  dplyr::select(-one_of("rhat", "nnet")) %>% 
  melt(id.vars="iter") %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value)) +
  geom_boxplot() +
  xlab("ML model") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14))
ggsave("../output/ml_comparison_ar1.pdf", g, width = 12, height = 8)
```

@@

tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
  ggplot(aes(x=var1, y=var2, colour=as.factor(chain))) +
  geom_jitter() +
  geom_smooth(method="lm", se=F)

tibble(var1=testing_data$V7, var2=testing_data$V15, chain=plda) %>% 
  ggplot(aes(x=var1, colour=as.factor(chain))) +
  stat_density(fill="none", position="identity")

hist(map_dbl(seq(1, 8, 1), function(i) cor(testing_data %>% filter(chain==i) %>% select(-chain) %>% as.matrix())[7, 15]))
```

Visualise between-chain correlation
```{r}
x <- split_data(x_ncp)
cors <- map(seq(1, dim(x)[2], 1), ~cor(x[, ., ]))
m_sd <- matrix(nrow = 19, ncol = 19)
for(i in 1:19)
  for(j in 1:19)
    m_sd[i, j] <- sd(map_dbl(seq(1, dim(x)[2], 1), ~cors[[.]][i, j]))
tmp <- which(m_sd == max(m_sd), arr.ind = TRUE)
a <- tmp[1,1]
b <- tmp[1,2]
m_sd[a, b]
colnames(r)[a] <- "a"
colnames(r)[b] <- "b"

g <- r %>% 
  ggplot(aes(x=a, y=b, colour=as.factor(chain))) +
  geom_jitter(alpha=0.1) +
  geom_smooth(method="lm", se=F) +
  scale_color_brewer("Chain", palette = "Dark2")
ggsave("../output/ml_comparison_8_schools_ncp_1.pdf", g, width = 12, height = 8)

g <- r %>% 
  select(chain, a, b) %>% 
  melt(id.vars="chain") %>% 
  ggplot(aes(x=value, colour=as.factor(chain), group=as.factor(chain))) +
  stat_density(fill="none", position = "identity") +
  facet_wrap(~variable)
ggsave("../output/ml_comparison_8_schools_ncp_2.pdf", g, width = 12, height = 8)

g <- r %>% 
  select(chain, a, b) %>% 
  melt(id.vars="chain") %>% 
  group_by(variable, chain) %>% 
  summarise(value=mean(value)) %>% 
  ggplot(aes(x=as.factor(chain), y=value)) +
  geom_point() +
  facet_wrap(~variable)
ggsave("../output/ml_comparison_8_schools_ncp_3.pdf", g, width = 12, height = 8)

g <- r %>% 
  select(chain, a, b) %>% 
  melt(id.vars="chain") %>% 
  group_by(variable, chain) %>% 
  summarise(value=quantile(value, 0.95)) %>% 
  ggplot(aes(x=as.factor(chain), y=value)) +
  geom_point() +
  facet_wrap(~variable)
ggsave("../output/ml_comparison_8_schools_ncp_4.pdf", g, width = 12, height = 8)
```

How R* varies with hyperparameters
```{r}
x <- x_ncp

f_run_gbm <- function(x, n.trees_, int.depth_){
  return(f_replicate_ml_simple(x, "gbm", expand.grid(interaction.depth=c(int.depth_), 
                             n.trees = n.trees_,
                             shrinkage=c(0.1),
                             n.minobsinnode=10)))
}

hypers <- tibble(
       n.trees=c(50, 50, 100, 200, 200, 200, 200, 400, 400, 1000),
       int.depth=c(3,5,3,3,5,7,9,5,9,11))
k <- 1
niter <- 20
m_results <- matrix(nrow = nrow(hypers) * niter, ncol=4)
for(i in 1:nrow(hypers)){
  for(j in 1:niter){
    tree <- hypers$n.trees[i]
    depth <- hypers$int.depth[i]
    temp <- f_run_gbm(x, tree, depth)
    m_results[k, ] <- c(tree, depth, j, temp)
    k <- k + 1
  }
}
colnames(m_results) <- c("n.tree", "int.depth", "iter", "r_star")
m_results <- m_results %>% 
  as.data.frame()
saveRDS(m_results, "../data/8_schools_gbm_hyper.rds")
m_results <- readRDS("../data/8_schools_gbm_hyper.rds")
g <- m_results %>% 
  group_by(n.tree, int.depth) %>% 
  summarise(lower=quantile(r_star, 0.25),
            middle=median(r_star),
            upper=quantile(r_star, 0.75)) %>% 
  ggplot(aes(x=as.factor(n.tree), y=middle, color=as.factor(int.depth))) +
  geom_pointrange(aes(ymin=lower, ymax=upper), position = position_dodge2(width=0.2)) +
  ylim(0.95, NA) +
  geom_hline(yintercept = 1, linetype=2) +
  scale_color_brewer("Int.depth", palette = "Dark2") +
  xlab("# tree") +
  ylab("R*")
ggsave("../output/ml_comparison_8_schools_hyper.pdf", g, width = 12, height = 8)
```

Try gbm function itself to optimise hyperparameters
```{r}
x <- x_ncp
x <- split_data(x)

rand_samples <- sample(1:nrow(r), 0.7 * nrow(r))
training_data <- r[rand_samples, ]
testing_data <- r[-rand_samples, ]

gbmFit <- train(chain ~ ., data = training_data, 
               method="gbm",
               tuneGrid = expand.grid(
       n.trees=c(50, 200, 2000),
       interaction.depth=c(3,11, 17),
       shrinkage=c(0.1),
             n.minobsinnode=c(10)))
plda <- predict(object=gbmFit, newdata=testing_data)
a_accuracy <- 
  tibble(predicted=plda, actual=testing_data$chain) %>%
  mutate(correct=if_else(predicted==actual, 1, 0)) %>% 
  summarise(mean(correct)) %>% 
  pull()
a_accuracy * n_distinct(training_data$chain)
varImp(gbmFit)
```

# Cauchy model
```{r}
fit_nom <- stan(file = 'cauchy_nom.stan', seed = 7878,
                refresh = 0)
x <- rstan::extract(fit_nom, permuted=F)

tunegrid_rf <- tibble(mtry = 1:2)
tunegrid_gbm <- expand.grid(interaction.depth=c(3), 
                             n.trees = 50,
                             shrinkage=c(0.1),
                             n.minobsinnode=10)
tunegrid_knn <- tibble(k = c(5, 10, 15, 20, 40))
tunegrid_svm <- tibble(C = c(0.25, 0.5, 0.75))
tunegrid_nnet <- tibble(size=c(2, 3, 4, 5), decay=c(0.5, 0.5, 0.5, 0.5))
tunegrid_multinom <- tibble(decay=c(0.1, 0.2, 0.5, 1))
niterates <- 10
m_res <- matrix(nrow=niterates, ncol = 6)
l_rf_hyper <- vector(length = niterates, mode="list")
l_knn_hyper <- vector(length = niterates, mode="list")
l_svm_hyper <- vector(length = niterates, mode="list")
l_nnet_hyper <- vector(length = niterates, mode="list")
l_multinom_hyper <- vector(length = niterates, mode="list")
for(i in 1:niterates){
  m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
  m_res[i, 2] <- max(f_replicate_rhat_simple(x))
  temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
  l_rf_hyper[[i]] <- temp$hyper
  m_res[i, 3] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
  l_knn_hyper[[i]] <- temp$hyper
  m_res[i, 4] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
  l_svm_hyper[[i]] <- temp$hyper
  m_res[i, 5] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "multinom", tunegrid_multinom)
  l_nnet_hyper[[i]] <- temp$hyper
  m_res[i, 6] <- temp$a_accuracy
}

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "multinom")
m_res1 <- m_res %>%
  as.data.frame() %>% 
  mutate(iter=seq_along(gbm))
saveRDS(m_res1, "../data/ml_comparison_cauchy_nom.rds")

g <- m_res1 %>% 
  dplyr::select(-one_of("rhat")) %>% 
  melt(id.vars="iter") %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value)) +
  geom_boxplot() +
  xlab("ML model") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14))
ggsave("../output/ml_comparison_cauchy_nom.pdf", g, width = 12, height = 8)

fit_alt1 <- stan(file = 'cauchy_alt_1.stan',
                 seed = 7878, refresh = 0)
x <- rstan::extract(fit_alt1, permuted=F)

m_res <- matrix(nrow=niterates, ncol = 6)
l_rf_hyper <- vector(length = niterates, mode="list")
l_knn_hyper <- vector(length = niterates, mode="list")
l_svm_hyper <- vector(length = niterates, mode="list")
l_nnet_hyper <- vector(length = niterates, mode="list")
l_multinom_hyper <- vector(length = niterates, mode="list")
for(i in 1:niterates){
  m_res[i, 1] <- f_replicate_ml_simple(x, "gbm", tunegrid_gbm)
  m_res[i, 2] <- max(f_replicate_rhat_simple(x))
  temp <- f_replicate_ml_simple_gridded(x, method_="rf", caretGrids_ = tunegrid_rf)
  l_rf_hyper[[i]] <- temp$hyper
  m_res[i, 3] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "knn", tunegrid_knn)
  l_knn_hyper[[i]] <- temp$hyper
  m_res[i, 4] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "svmLinear", tunegrid_svm)
  l_svm_hyper[[i]] <- temp$hyper
  m_res[i, 5] <- temp$a_accuracy
  temp <- f_replicate_ml_simple_gridded(x, "multinom", tunegrid_multinom)
  l_multinom_hyper[[i]] <- temp$hyper
  m_res[i, 6] <- temp$a_accuracy
}

colnames(m_res) <- c("gbm", "rhat", "rf", "knn", "svm", "multinom")
m_res1 <- m_res %>%
  as.data.frame() %>% 
  mutate(iter=seq_along(gbm))
saveRDS(m_res1, "../data/ml_comparison_cauchy_alt.rds")
m_res1 <- readRDS("../data/ml_comparison_cauchy_alt.rds")
m_res2 <- readRDS("../data/ml_comparison_cauchy_nom.rds") %>% 
  mutate(type="nominal")

g <- m_res1 %>% 
  mutate(type="alternative") %>% 
  bind_rows(m_res2) %>% 
  dplyr::select(-one_of("rhat")) %>% 
  melt(id.vars=c("iter", "type")) %>% 
  mutate(value=as.numeric(as.character(value))) %>% 
  ggplot(aes(x=fct_reorder(as.factor(variable), value), y=value, fill=as.factor(type))) +
  geom_boxplot() +
  xlab("ML model") +
  ylab("R*") +
  theme(text=element_text(colour="black", size=14)) +
  scale_fill_brewer("Parameterisation", palette = "Dark2") +
  scale_y_continuous(limits=c(0.8, NA)) +
  geom_hline(yintercept = 1, linetype=2)
ggsave("../output/ml_comparison_cauchy.pdf", g, width = 12, height = 8)
```

### 2020-10-09 write_up/arxiv_submission_aug_2020/ms_ba_marked.tex
set up infrastructure for hyperparameter runs and ml comparison

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about \textcolor{red}{its performance in finite time}. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution. Here, we introduce a new method for diagnosing convergence based on \textcolor{red}{how well} a machine learning classifier model can successfully discriminate the individual chains. We call this convergence measure $R^*$. In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; \textcolor{red}{instead it uses} all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches. Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. The method is straightforward to implement, robust to GBM hyperparameter choice, and could be a complementary additional check on MCMC convergence for applied analyses.
\end{abstract}

@@

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it. \textcolor{red}{Rephrased, it should not be possible to predict which draws come from which chain}. In this vein, we use a machine learning (ML) classifier to measure convergence. Specifically, we train a ML classifier to predict the chain that generated each observation. By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics. Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution. \textcolor{red}{However, the ML classifier we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.} For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer. It is also insensitive to GBM's hyperparameters and provides a measure of convergence robust to various Markov chain pathologies. In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

\subsubsection{Robustness to hyperparameter choice}\label{sec:heterogeneity_robustness}
The performance of GBM, like all ML methods, depends on its hyperparameters. To investigate the robustness of $R^*$ to GBM hyperparameter choice, we performed a sensitivity analysis. In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig. \ref{fig:ar1}. The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200). The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig. \ref{fig:ar1_sensitivity}. These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper). This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.

@@

>>>>

\begin{abstract}
	Markov chain Monte Carlo (MCMC) has transformed Bayesian model inference over the past three decades: mainly because of this, Bayesian inference is now a workhorse of applied scientists. Under general conditions, MCMC sampling converges asymptotically to the posterior distribution, but this provides no guarantees about \textcolor{red}{its performance in finite time}. The predominant method for monitoring convergence is to run multiple chains and monitor individual chains' characteristics and compare these to the population as a whole: if within-chain and between-chain summaries are comparable, then this is taken to indicate that the chains have converged to a common stationary distribution. Here, we introduce a new method for diagnosing convergence based on \textcolor{red}{how well} a machine learning classifier model can successfully discriminate the individual chains. We call this convergence measure $R^*$. In contrast to the predominant $\widehat{R}$, $R^*$ is a single statistic across all parameters that indicates lack of mixing, although individual variables' importance for this metric can also be determined. Additionally, $R^*$ is not based on any single characteristic of the sampling distribution; \textcolor{red}{instead it uses} all the information in the chain, including that given by the joint sampling distribution, which is currently largely overlooked by existing approaches. Since our choice of machine learning classifier, a gradient-boosted regression trees model (GBM), provides uncertainty in predictions, as a byproduct, we obtain uncertainty in $R^*$. \textcolor{red}{The method is straightforward to implement and could be a complementary additional check on MCMC convergence for applied analyses.}
\end{abstract}

@@

Here, we introduce $R^*$, a new convergence metric. This statistic is built on the intuition that, if chains are mixed, it should not be possible to discern from a draw's value the chain that generated it. \textcolor{red}{Rephrased, it should not be possible to predict which draws come from which chain}. In this vein, we use a machine learning (ML) classifier to measure convergence. Specifically, we train a ML classifier to predict the chain that generated each observation. By evaluating the performance of the classifier on a held-out test set, this provides a new convergence metric. To maximise predictive accuracy, our chosen ML classifier naturally exploits differences in the full joint distributions between chains, which means it's sensitive to variations across the joint distribution of target model dimensions unlike most existent convergence diagnostics. Our statistic, unlike its $\widehat{R}$ cousins, is scalar valued for multivariate distributions: one model provides a single $R^*$, whereas $\widehat{R}$ has separate values for each univariate marginal distribution. \textcolor{red}{However, the ML classifier we use can straightforwardly be interrogated to estimate which parameters were most important for generating predictive accuracy.} For our ML classifier, we use gradient-boosted regression trees \citep{friedman2001greedy,greenwell2019package} (``GBM''), since these are known to perform well for the types of tabular data that our problem presents \citep{chollet2018}. For the types of problem we test, $R^*$ calculation is of a speed comparable to some of the newer $\widehat{R}$ measures calculated (typically $\mathcal{O}$(seconds) to calculate), although for models with 10,000s of parameters and many iterations, the time taken is longer. \textcolor{red}{The statistic depends on the hyperparameters of GBM, although even for ``default'' settings, $R^*$ provides a stringent test of accuracy, giving a reliable measure of various Markov chain pathologies.} In addition, since GBMs can output predicted class probabilities, we obtain uncertainty measures for $R^*$, which we find provides a useful summary of MCMC convergence. $R^*$ can straightforwardly be incorporated into existing software libraries to provide a complementary convergence metric alongside more established measures.

@@

@@

\color{red}
\subsection{Hyperparameter sensitivity}\label{sec:hyperparameters}
The performance of GBM, like all ML methods, depends on its hyperparameters. In GBMs, a regression function is approximated by additive components, where each of those corresponds to a tree \cite{greenwell2019package}. The number of such regression trees is thus a hyperparameter of the model. Each regression tree can split amongst any integer number of variables (the \textit{interaction depth}). In this section, we investigate the sensitivity of $R^*$ to these hyperparameters across a range of examples.

\subsubsection{Autoregressive example}\label{sec:heterogeneity_robustness}
In this, we fit GBMs with six different sets of hyperparameters to replicates consisting of the same data used to generate Fig. \ref{fig:ar1}. The hyperparameters we varied were the interaction depth (3, 5 or 7) and the number of trees (50, 100 or 200). The resultant $R^*$ point estimates across 200 replicates for each hyperparameter setting are shown in Fig. \ref{fig:ar1_sensitivity}. These show that hyperparameter choice affects $R^*$, with the highest median values obtained for an interaction depth of 3 with 50 trees (our default across all the examples we consider in this paper). This analysis does suggest, however, that this effect is relatively minor: indeed, across the various hyperparameter sets considered, the percentage of replicates where $R^*<1$ did not strongly vary.

\subsubsection{Eight schools: non-centered parameterisation}

\subsubsection{Cauchy model: alternative paramerisation}

\subsubsection{Multivariate normal: 250-dimensional model}

\subsection{ML model sensitivity}\label{sec:ml_model}

\color{black}
